Joint Likelihood Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2023-05-10 src/tex/ms.tex
[showyourwork] overleaf sync

    First astronomical x-ray and \gammaray data data is rare, expensive to obtain and there is no ground truth, because we can only observe the sky with our given instruments. To circumvent this one could rely on transfer learning and simulated data. For example radio or optical images could be forward folded with a known PSF model and be degraded with Poisson noise to generate training data. However astronomical images still have a large dynamic range and show a high diversity in morphological structure, which would still require a large amount of training data. However this approach has been successfully applied for specific classes of sources, such as images of Galaxies in the Sloan Digital Sky Survey (SDSS) \citep{Schawinski2017} or as a post processing step of results obtained with regularised RL \citep{Akhaury2022}.

    In addition to the challenge of obtaining training data, it is known that the PSF of instruments such as \chandra is highly variable. It varies with observation conditions and depends e.g. on energy as well as offset angle from the pointing direction. To cover the whole parameter space of the PSF one would need to train either a network for each of the PSF models, or define a very large network, which would require a very large computational effort.

    Some of the challenges of deblurring with CNN are due to the convolutional nature of the networks. In a single layer, linear approximation, the network would need to learn the inverse filter of the PSF, corresponding to the well known \textit{Wiener filter} solution of the problem. This requires large convolution kernels or deep network architectures, which require in turn large amounts of training data. This challenge was described early by \cite{Li2014} who introduced a baseline architecture for a deep CNN for image deblurring.

    One last important challenge for deep learning based methods is obtaining uncertainty estimates for the reconstructed image from CNNs. In the limit of Poisson noise it is desirable to trace the full a-postiori likelihood, which is not possible for most network architectures, with the exception of for example normalising flows.
    This problem could be adresses with Monte Carlo base methods, such as bootstrapping, further increasing the demand on training data and computation ressources.

@@

    To account for the possibility of different datasets having a different number of pixels we divide by $N$ (\todo{Does this need more justification...?}).
    By summing over all observations we finally get the joint log-likelihood of measuring a set of counts images $\mathbf{D}$ given an unknown flux image $\mathbf{x}$:
    %
    \begin{equation}
        \label{eq:joint}
        \mathcal{L}\left( \mathbf{D} | \mathbf{x} \right) = \sum_{j=1}^J \mathcal{C}\left( \mathbf{D_j} | \mathbf{x} \right)
    \end{equation}

@@

    \begin{figure}
        \script{patches.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/patches.pdf}
            \caption{
                Grid of overlapping quadratic patches of size $8\times8$ pixels on top of an example image of the size $128 \times 68$ pixels. The overlap size was chosen to be $2$ pixels, which results in 50 patches in total. The patches are shown translucent, such that the light \enquote{stripes} result from two neighboring overlapping patches, while the small, even lighter, squares result from four patches overlapping at the their corners.
            }
            \label{fig:patches}
        \end{centering}
    \end{figure}

@@

    \subsubsection{JWST Cassiopeia A}
    The James Webb Space Telescope (JWST) is a large, infrared space telescope launched in late 2021. It delivers image data a very high spatial resolution as well as excellent signal to noise ratio. Both of these properties make JWST data an excellent choice for extracting patches for training a GMM. To limit the data reduction effort we relied on a promotional image of the SNR \textit{Cassiopeia A}\footnote{The image was downloaded from \url{https://webbtelescope.org/contents/media/images/2023/121/01GWQBBY77MHGFV3M3N63KDCEJ}}.
    We first converted the image to gray scale using a perceived brightness model \todo{reference skimage...}. Then we split the image into $approx 2$ million patches and learned a GMM with 128 mixture components.

@@

    \begin{figure}
        \script{gmm-eigen-images.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/gmm-eigen-images.pdf}
            \caption{
                Eigenimages of the covariance matrices of the 32 components of the GMM learned from the NRAO jets. \todo{Provide other examples...?!?}
            }
            \label{fig:gmm-eigen-images}
        \end{centering}
    \end{figure}

@@

    The ground truth and corresponding expected counts for the combination of scenarios are shown in Figure~\ref{fig:scenarios}, and the details of the locations and brightness of each component is listed in Table~\ref{tab:scenarios}.

@@

    Constant background levels of $\lambda_{Bkg}= 0.001, 0.01 \textrm{and} 0.1 \textrm{counts/pixel}$.

@@

    \subsubsection{Comparing Signal-to-Noise Scenarios}
    %
    \begin{figure*}
        \script{comparison-signal-noise.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/comparison-signal-noise.pdf}
            \caption{
                Comparison of the \jolideco method with the GLEAM v0.1 prior for different background levels and signal strength represented by scenario D1-5. The images are grouped by background levels with each column group showing the counts data in the left column and the reconstruction in the right column. The rows show the scenarios D1-5 as described in Section~\ref{subsec:test-datasets}. The background levels correspond to \qty[mode = text]{0.001}{cts / pixel}, \qty[mode = text]{0.01}{cts / pixel} and \qty[mode = text]{0.1}{cts / pixel}. The complete results for the simulations and additional information are available at \url{https://jolideco.github.io/jolideco-comparison}. \todo{the background handling does not yet seem right...maybe we should let the background norm free}
            }
            \label{fig:comparison-signal-noise}
        \end{centering}
    \end{figure*}
    %
    As a second experiment we explored the dependency of the results on the background level and signal strength. Figure~\ref{fig:comparison-signal-noise} shows the results for the \jolideco method with the GLEAM prior in the scenario D1-5. The background level was varied between

    As expected the reconstruction quality improves with signal strength. This includes both the variation of the background as well as signal strength. \todo{sensitivity scales as approx signal / sqrt(Bkg)? } Visually the method converges to the ground truth for very high signal-to-noise ratio. Once the signal strength becomes very large there is no dependency of the result on the background level anymore (especially valid for scenario D5)

@@

    \begin{figure*}
        \script{example-chandra.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/example-chandra.pdf}
            \caption{
               The \jolideco method applied to \chandra data of the SNR \textit{1E~0102.2–7219}. The left large image shows the summed counts from 25 observations. The right large image  shows the result from the \jolideco deconvolution process. The flux is given as counts relative to the reference observation with ID=1308. The panels \textit{Zoom A}, \textit{Zoom B} and \textit{Zoom C} show a zoomed-in version of three example regions in the SNR, marked with the white rectangles in the large images. The colormap and scale used in the zoom insets is the same as used for the larger images.
            }
            \label{fig:example-chandra}
        \end{centering}
    \end{figure*}

    As a first application example for real data we choose a set of short exposure observations of the SNR \textit{1E~0102.2–7219} by the \chandra telescope and combine them into a single deep observation using \jolideco and the patch prior learned from JWST data. In total we chose 25 observations\footnote{Same list of observation IDs as for \url{https://chandra.si.edu/photo/2009/e0102/}} taken with with the ACIS instrument and no grating. For \chandra the instrument response varies with the observation offset and energy \todo{reference}. With \jolideco we can handle  PSF for each observation. We assume a spatially constant background of \qty[mode = text]{1e-5}{cts/pixel} but allow the norm to vary later in the reconstruction process. We assume the exposure to be spatially constant and express its absolute value for each observation as the ratio of observation time with respect to a reference observation. As reference we chose the observation with id 1308.

    We reduced the data using \texttt{ciao} v4.15 and a custom \texttt{snakemake} workflow (see also Section~\ref{sec:reproducibility} on reproducibility). The PSF were simulated for each observation independently using the tool \texttt{marx} \todo{reference} and the \texttt{simulate\_psf} script. We selected the energy range from \qty[mode = text]{0.5}{keV} to  \qty[mode = text]{7}{keV} and a bins size of 0.25 native bins, corresponding to a bin size of \qty[mode = text]{0.001}{$\deg$/pixel} \todo{How large is the Chandra PSF compared to this?}. As reference spectrum for the PSF simulation we assumed the model derived by \cite{Plucinsky2017}. We converted the spectral from XSPEC to sherpa script using the helper script \texttt{convert\_xspec\_script}. As reported in \cite{Xi2019} the effect of pileup, even for the brightest regions is small (<5\%), so we did not include the effect of pileup in our PSF simulations.

    For \jolideco we used the GMM prior learned on the JWST Cas A image. As there is a lot of statistics in the image we used and oversampling factor of 2. For each observation we allowed the background level and position calibration to vary. As reference for the positional shift we used the observation with the ID 1308. 
    Figure~\ref{fig:example-chandra} shows the final result. \jolideco clearly improves both the angular resolution as well as reduces the noise of the counts image.

@@

    Checking the residuals per observation (not shown here), we noticed that the assumption of the time independence of the source as well the instrument response is not completely fulfilled in this case. We saw the detection efficiency of \chandra changed between the observations as well as sub-structure in the source expands over time. Both effects are previously known have been reported e.g. in \cite{Xi2019} and \cite{}. Despite the clear improvement in spatial resolution the proper motion of structures in the image might limit the quality of the result. The full analysis example is available at \url{https://github.com/jolideco/jolideco-chandra-examples}

>>>>

    Firstly, astronomical x-ray and \gammaray data data is rare, expensive to obtain and there is no ground truth, because we can only observe the sky with our given instruments. To circumvent this one could rely on transfer learning and simulated data. For example radio or optical images could be forward folded with a known PSF model and be degraded with Poisson noise to generate training data. However astronomical images still have a large dynamic range and show a high diversity in morphological structure, which would still require a large amount of training data. However this approach has been successfully applied for specific classes of sources, such as images of Galaxies in the Sloan Digital Sky Survey (SDSS) \citep{Schawinski2017} or as a post processing step of results obtained with regularised RL \citep{Akhaury2022}. \cite{Sweere2022} used images from
    the large scale structure simulations with promising results.

    In addition to the challenge of obtaining training data, it is known that the PSF of instruments such as \chandra is highly variable. It varies with observation conditions and depends e.g. on energy as well as offset angle from the pointing direction. To cover the whole parameter space of the PSF one would need to train either a network for each of the PSF models, or define a very large network, which allows to represent the the varying PSF as well.

    Some of the challenges of deblurring with CNN are due to the convolutional nature of the networks. In a single layer, linear approximation, the network would need to learn the inverse filter of the PSF, corresponding to the well known \textit{Wiener filter} solution of the problem. This requires either large convolution kernels or deep network architectures, which require in turn large amounts of training data. This challenge was described early by \cite{Li2014} who introduced a baseline architecture for a deep CNN for image deblurring.

    One last important challenge for deep learning based methods is obtaining uncertainty estimates for the reconstructed image from CNNs. In the limit of Poisson noise it is desirable to trace the full a-postiori likelihood, which is not possible for most network architectures, with the exception of for example normalising flows. This problem could be addressed with Monte Carlo based methods, such as bootstrapping, which further increases the demand on training data and computational resources. For these reasons we decided to choose an approach more close to \enquote{classical} statistical modelling techniques. However we will include elements from ML to represent the prior information on images.

@@

    To account for the possibility of different datasets having a different number of pixels we divide by $N$ (\todo{Does this need more justification...is it reasonable?}).
    By summing over all observations we finally get the joint log-likelihood of measuring a set of counts images $\mathbf{D}$ given an unknown flux image $\mathbf{x}$:
    %
    \begin{equation}
        \label{eq:joint}
        \mathcal{L}\left( \mathbf{D} | \mathbf{x} \right) = \sum_{j=1}^J \mathcal{C}\left( \mathbf{D_j} | \mathbf{x} \right)
    \end{equation}

@@

    \begin{figure}
        \script{patches.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/patches.pdf}
            \caption{
                Grid of overlapping quadratic patches of size $8\times8$ pixels on top of an example image of the size $128 \times 68$ pixels. The overlap size was chosen to be $2$ pixels, which results in 50 patches in total. The patches are shown translucent, such that the light \enquote{stripes} result from two neighboring overlapping patches, while the small, even lighter squares result from four patches overlapping at the their corners.
            }
            \label{fig:patches}
        \end{centering}
    \end{figure}

@@

    \subsubsection{JWST Cassiopeia A}
    The James Webb Space Telescope (JWST) is a large, infrared space telescope launched in late 2021. It delivers image data a very high spatial resolution as well as excellent signal to noise ratio. Both of these properties make JWST data an excellent choice for extracting patches for training a GMM. To limit the data reduction effort we relied on a promotional image of the SNR \textit{Cassiopeia A}\footnote{We downloaded the full resolution (4k)  PNG image from \url{https://webbtelescope.org/contents/media/images/2023/121/01GWQBBY77MHGFV3M3N63KDCEJ}}.
    We first converted the image to gray scale using a perceived brightness model \todo{reference skimage...}. Then we split the image into $approx 2$ million patches and learned a GMM with 128 mixture components.

@@

    \begin{figure}
        \script{gmm-eigen-images.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/gmm-eigen-images.pdf}
            \caption{
                Weighted sum of Eigenimages of the covariance matrices of the first $32$ components of the GMM learned from the JWST Cas A image. The color scale varies between the images, so their contrast is different. 
            }
            \label{fig:gmm-eigen-images}
        \end{centering}
    \end{figure}

@@

    The ground truth and corresponding expected counts for all combinations of source and instrument scenarios are shown in Figure~\ref{fig:scenarios}. The details of the locations and brightness of each component is listed in Table~\ref{tab:scenarios}. In addition we also assume a constant background with varying levels of $\lambda_{Bkg}= 0.001, 0.01$ and \qty[mode = text]{0.1}{counts/pixel}.

@@

@@

    \subsubsection{Comparing Signal-to-Noise Scenarios}
    %
    \begin{figure*}
        \script{comparison-signal-noise.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/comparison-signal-noise.pdf}
            \caption{
                Comparison of the \jolideco method with the GLEAM v0.1 prior for different background levels and signal strength represented by scenario D1-5. The images are grouped by background levels with each column group showing the counts data in the left column and the reconstruction in the right column. The rows show the scenarios D1-5 as described in Section~\ref{subsec:test-datasets}. The background levels correspond to \qty[mode = text]{0.001}{cts / pixel}, \qty[mode = text]{0.01}{cts / pixel} and \qty[mode = text]{0.1}{cts / pixel}. The complete results for the simulations and additional information are available at \url{https://jolideco.github.io/jolideco-comparison}.
            }
            \label{fig:comparison-signal-noise}
        \end{centering}
    \end{figure*}
    %
    As a second experiment we explored the dependency of the results on the background level and signal strength. Figure~\ref{fig:comparison-signal-noise} shows the results for the \jolideco method with the GLEAM prior in the scenario D1-5. The background level was varied between 
    $\lambda_{Bkg}= 0.001, 0.01$ and \qty[mode = text]{0.1}{counts/pixel}.

    As expected the reconstruction quality improves with signal strength. This includes both the variation of the background as well as signal strength. \todo{sensitivity scales as approx signal / sqrt(Bkg)? so what what is the implication? } Visually the method converges to the ground truth for very high signal-to-noise ratio. Once the signal strength becomes very large there is no dependency of the result on the background level anymore. This especially valid for scenario D5

@@

    \begin{figure*}
        \script{example-chandra.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/example-chandra.pdf}
            \caption{
               Result of the \jolideco method applied to \chandra data of the SNR \textit{1E~0102.2–7219}. The left large image shows the summed counts from 25 observations. The right large image  shows the result from the \jolideco deconvolution process. The flux is given as counts relative to the reference observation with ID=1308. The panels \textit{Zoom A}, \textit{Zoom B} and \textit{Zoom C} show a zoomed-in version of three example regions in the SNR, marked with the white rectangles in the large images. The colormap and scale used in the zoom insets is the same as used for the larger images.
            }
            \label{fig:example-chandra}
        \end{centering}
    \end{figure*}

    As a first application example for real data we choose a set of short exposure observations of the SNR \textit{1E~0102.2–7219} by the \chandra telescope and combine them into a single deep observation using \jolideco and the patch prior learned from JWST data. In total we chose 25 observations\footnote{Same list of observation IDs as for \url{https://chandra.si.edu/photo/2009/e0102/}} taken with  the ACIS instrument and no grating.

    We reduced the data using \texttt{ciao} v4.15 and a custom \texttt{snakemake} workflow (see also Section~\ref{sec:reproducibility} on reproducibility). For the counts image we selected the energy range from \qty[mode = text]{0.5}{keV} to  \qty[mode = text]{7}{keV} and a pixel size of $0.25$ of the native pixel size of \chandra. This corresponds to a pixel size of \qty[mode = text]{0.001}{$\deg$/pixel} \todo{How large is the Chandra PSF compared to this?}.

    \chandra's instrument response varies with the observation offset and energy \todo{reference}, so we simulated the PSF for each observation independently using the tool \texttt{marx} \todo{reference} and the \texttt{simulate\_psf} script. As reference spectrum for the PSF simulation we assumed the model derived by \cite{Plucinsky2017}. We converted the spectral from XSPEC to sherpa script using the helper script \texttt{convert\_xspec\_script}. As reported in \cite{Xi2019} the effect of pileup, even for the brightest regions is small ($<5\%$), so we did not include the effect of pileup in our PSF simulations.

    We assume a spatially constant background of \qty[mode = text]{1e-5}{cts/pixel} but allow the norm to vary later in the reconstruction process. We assume the exposure to be spatially constant and express its absolute value for each observation as the ratio of observation time with respect to a reference observation. As reference we chose the observation with id 1308.

    For \jolideco we used the GMM prior learned on the JWST Cas A image. As there is a lot of statistics in the image we used and oversampling factor of 2. For each observation we allowed the background level and position calibration to vary. As reference for the positional shift we also used the observation with the ID 1308.

    Figure~\ref{fig:example-chandra} shows the final result. \jolideco clearly improves both the angular resolution as well as reduces the noise of the counts image.

@@

    Checking the residual images per observation (not shown here), we noticed that the assumption of the time independence of the source as well the instrument response is not completely fulfilled in this case. We saw the detection efficiency of \chandra changed between the observations as well as that sub-structure in the remnant expands over time. Both effects are previously known and have been reported e.g. in \cite{Xi2019} and \cite{}. Despite the clear improvement in spatial resolution the proper motion of structures in the image might limit the quality of the result. The full analysis example is available at \url{https://github.com/jolideco/jolideco-chandra-examples}
